
\section{Introduction}Over the last several decades we have observed the development of a 
large collection of specification and modeling languages and 
associated methodologies, and tools. Their purpose is to 
support modeling of requirements and high-level designs before 
programming is initiated. Agile approaches advocate to avoid 
explicit modeling entirely and suggest to go directly to coding. 
Other approaches advocate avoiding manual “coding” in a programming 
language entirely and suggest instead the generation of code 
directly from the models. This way modeling languages replace 
programming languages.  We can divide these modeling languages into 
formal specification languages (formal methods), usually focusing 
on textual languages based on mathematical logic and set theory, 
and associated proof tools (theorem provers, model checkers, etc.), 
and on the other hand model-based engineering languages (UML, 
SysML, Modelica, Mathematica, …), focusing more on design, code 
generation and simulation. Many of these modeling languages have 
similarities with programming languages.In parallel, and frankly seemingly independent, we have seen the 
development of numerous new programming languages. Few languages 
have had the success of C, which still today is the main 
programming language for embedded systems. The success is so 
outstanding that nearly no progress wrt. praxis has been made in 
this domain (embedded programming) since the 1970ties.  In 
application programming a collection of new languages came ago such 
as Ada, C++, Eiffel etc. At the same time we have seen several 
high-level languages appear targeting the softer side of software 
engineering (such as web-programming, user interfaces, scripting), 
including languages such as Java, JavaScript, Ruby, Python and 
Scala.  More academic languages include Haskell and the ML family, 
including OCaml.In this white paper we will give a brief overview of some of the, 
in our view, important developments in modeling and programming 
languages. Subsequently we will propose en effort to develop a 
specification, design and implementation language that combines 
modeling and programming.