
\section{Observed concepts in programming}Several new programming languages have emerged over the last 
decade, which include abstraction mechanisms known from the formal 
specification languages mentioned above. Such languages include 
Eiffel, Java, Python, Scala, Fortress,  C\#, Spec\#, F\#,  Dafny, D, 
RUST, Swift, Go, Agda, and SPARK.  Some languages support design-
by-contract with pre-post conditions, and in some cases with 
invariants. These languages  include for example Eiffel, Spec\#, 
Dafny, SPARK, and to some limited extent Scala. Java supports 
contracts through JML, which, however, is not integrated with Java, 
but an add-on comment language (JML specifications are comments in 
a Java program). Most of the languages above support abstract 
collections such as sets, lists and maps. It is interesting to 
observe that SUN’s Fortress language (which unfortunately was not 
finished by the designers) supports a mathematical notation for 
collections very similar to VDM. The Dafny language is interesting 
since it is developed specifically with specification and 
verification in mind.A trend on the rise is likely the combination of object oriented 
and functional programming, as seen in perhaps most prominently 
Scala, but also in the earlier Python, and now in Java which got 
closures in version 1.8.  Ocaml is a similar earlier attempt to 
integrate object oriented and functional programming, although in a 
layered manner, and not integrated with the standard module system. 
Some interesting new directions of research include dependent types 
as found in Agda (to some extent related to predicate subtypes in 
VDM) and session types. Session types are temporal patterns that 
can be checked at compile time. They are much related to temporal 
logic as used within the formal methods community to express 
properties of concurrent programs. At the same time there are also 
attempts to make more conservative moves away from C, but without 
losing too much efficiency. Examples include the languages D and 
RUST. However, as stated earlier, C has an impressive staying power, and these attempts have not yet made the highway.

